---
title: "Tibble Joins: A Tale of Two Tibbles"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(dplyr)

gradethis_setup()
knitr::opts_chunk$set(echo = FALSE)
```

# Tibble Joins: A Tale of Two Tibbles

Today's topic is on operations with two or more tibbles. These operations are 
are used to combine two different data tables, compare two different data tables,
verify or filter data in one data table against another, etc.


## Worksheet

You can find a worksheet template for today [here](https://raw.githubusercontent.com/USF-Psych-DataSci/Classroom/master/tutorials/s07_tibble-join-exercise.Rmd).

```{r}
fam_data <- read.csv("https://raw.githubusercontent.com/wesley4546/teach-r-temp/master/data/family_data.csv")
fam_long_data <- read.csv("https://raw.githubusercontent.com/wesley4546/teach-r-temp/master/data/family_long_data.csv")
```

## Requirements

You will need data from Joey Bernhardt's `singer` R package for this exercise. 

You can download the singer data from the `USF-Psych-DataSci/Classroom` repo:

```{r}
songs <- read.csv("https://github.com/USF-Psych-DataSci/Classroom/raw/master/data/singer/songs.csv")
locations <- read.csv("https://github.com/USF-Psych-DataSci/Classroom/raw/master/data/singer/loc.csv")
```

If you want, you could instead install the `singer` package itself. To do that,
you'll need to install `devtools`. Running this code in your console should do 
the trick:

```
install.packages("devtools")
devtools::install_github("JoeyBernhardt/singer")
```

Load required packages:

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(singer)
knitr::opts_chunk$set(fig.width=4, fig.height=3, warning = FALSE, fig.align = "center")
```

<!-- The following chunk allows errors when knitting -->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE)
```

## Exercise 1: `singer`

The package `singer` comes with two smallish data frames about songs. 
Let's take a look at them (after minor modifications by renaming and shuffling):

```{r echo=TRUE}
(time <- as_tibble(songs) %>% 
   rename(song = title))
```

```{r echo=TRUE}
(album <- as_tibble(locations) %>% 
   select(title, everything()) %>% 
   rename(album = release,
          song  = title))
```


1. We really care about the songs in `time`. But, for which of those songs do we 
   know its corresponding album?

```{r inner_join_exercise, exercise = TRUE}
time %>% 
  FILL_THIS_IN(album, by = FILL_THIS_IN)
```

```{r inner_join_exercise-solution}
time %>% 
  inner_join(album, by = c("song","artist_name"))
```

```{r inner_join_exercise-check, echo = FALSE}
grade_code()
```

2. Go ahead and add the corresponding albums to the `time` tibble, being sure to 
   preserve rows even if album info is not readily available.

```{r left_join_exercise, exercise = TRUE}
time %>% 
  FILL_THIS_IN(album, by = FILL_THIS_IN)
```

```{r left_join_exercise-solution}
time %>% 
  left_join(album, by = c("song","artist_name"))
```

```{r left_join_exercise-check, echo = FALSE}
grade_code()
```

3. Which songs do we have "year", but not album info?

```{r anti_join_exercise, exercise = TRUE}
time %>% 
  FILL_THIS_IN(album, by = "song")
```

```{r anti_join_exercise-solution}
time %>% 
  anti_join(album, by = "song")
```

```{r anti_join_exercise-check, echo = FALSE}
grade_code()
```

4. Which artists are in `time`, but not in `album`?

```{r anti_join_two_exercise, exercise = TRUE}
time %>% 
  anti_join(album, by = "FILL_THIS_IN")
```

```{r anti_join_two_exercise-solution}
time %>% 
  anti_join(album, by = "artist_name")
```

```{r anti_join_two_exercise-check, echo = FALSE}
grade_code()
```

5. You've come across these two tibbles, and just wish all the info was 
   available in one tibble. What would you do?

```{r full_joing_exercise, exercise = TRUE}
FILL_THIS_IN %>% 
  FILL_THIS_IN(FILL_THIS_IN, by = "song")
```

```{r full_joing_exercise-solution}
album %>% 
  full_join(select(album, -artist_name), by = "song")
```

```{r full_joing_exercise-check, echo = FALSE}
album %>% 
  full_join(select(album, -artist_name), by = "song")
```

## Exercise 2: LOTR

Load in three tibbles of data on the Lord of the Rings:

```{r include=FALSE}
fell <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Fellowship_Of_The_Ring.csv")
ttow <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Two_Towers.csv")
retk <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Return_Of_The_King.csv")
```

```{r echo=TRUE}
fell 
ttow 
retk 
```


1. Combine these into a single tibble.

```{r bing_rows_exercise, exercise = TRUE}
FILL_THIS_IN(fell, FILL_THIS_IN)
```

```{r bing_rows_exercise-solution}
bind_rows(fell,ttow,retk)
```

```{r bing_rows_exercise-check, echo = FALSE}
grade_code()
```

2. Which races are present in "The Fellowship of the Ring" (`fell`), but not in 
   any of the other ones?

```{r anti_join_exercise_lotr, exercise = TRUE}
fell %>% 
  anti_join(ttow, by = "Race") %>% 
  FILL_THIS_IN(FILL_THIS_IN, by = "Race")
```

```{r anti_join_exercise_lotr-solution}
fell %>% 
  anti_join(ttow, by = "Race") %>% 
  anti_join(retk, by = "Race")
```

```{r anti_join_exercise_lotr-check, echo = FALSE}
grade_code()
```

## Exercise 3: Set Operations

Let's use three set functions: `intersect`, `union` and `setdiff`. We'll work 
with two toy tibbles named `y` and `z`, similar to Data Wrangling Cheatsheet

```{r}
(y <-  tibble(x1 = LETTERS[1:3], x2 = 1:3))
```

```{r}
(z <- tibble(x1 = c("B", "C", "D"), x2 = 2:4))
```

1. Rows that appear in both `y` and `z`

```{r intersect_exercise, exercise = TRUE}
FILL_THIS_IN(y, z)
```

```{r intersect_exercise-solution}
intersect(y, z)
```

```{r intersect_exercise-check, echo = FALSE}
grade_code()
```

2. You collected the data in `y` on Day 1, and `z` in Day 2. 
   Make a data set to reflect that.

```{r union_exercise, exercise = TRUE}
FILL_THIS_IN(
  mutate(y, day = "Day 1"),
  mutate(z, day = "Day 2")
)
```

```{r union_exercise-solution}
union(
  mutate(y, day = "Day 1"),
  mutate(z, day = "Day 2")
)
```

```{r union_exercise-check, echo = FALSE}
grade_code()
```

3. The rows contained in `z` are bad! Remove those rows from `y`.

```{r setdiff_exercise, exercise = TRUE}
FILL_THIS_IN(FILL_THIS_IN, FILL_THIS_IN)
```

```{r setdiff_exercise-solution}
setdiff(y, z)
```

```{r setdiff_exercise-check, echo=FALSE}
grade_code()
```


## Resources

- [Jenny Bryan's join cheatsheet](https://stat545.com/join-cheatsheet.html)
- The `dplyr` "two-table verbs" [vignette](https://cran.r-project.org/web/packages/dplyr/vignettes/two-table.html)
- [Relational Data chapter](https://r4ds.had.co.nz/relational-data.html) in "R for Data Science".
- [dplyr cheatsheet](https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)

For an overview of operations involving multiple tibbles, check out Jenny Bryan's 
[Chapter 14](https://stat545.com/multiple-tibbles.html) in stat545.com.

For more activities, see the matrials from  
[Rashedul Islam](https://github.com/Rashedul/stat545_guest_lecture).


## Join Functions

Often, we need to work with data living in more than one table. There are four 
main types of operations that can be done with two tables (as elaborated in 
[r4ds Chapter 13 Introduction](https://r4ds.had.co.nz/relational-data.html#introduction-7)):

- [__Mutating joins__](https://r4ds.had.co.nz/relational-data.html#mutating-joins) 
  add new columns to the "original" tibble.
- [__Filtering joins__](https://r4ds.had.co.nz/relational-data.html#filtering-joins) 
  filter the "original" tibble's rows.
- [__Set operations__](https://r4ds.had.co.nz/relational-data.html#set-operations) 
  work as if each row is an element in a set. 
- __Binding__ stacks tables on top of or beside each other, with 
  `bind_rows()` and `bind_cols()`.

Let's navigate to each of these three links, which lead to the relevant r4ds 
chapters, and go through the concepts there. These have excellent visuals to 
explain what's going on.

We will also look at the visuals of these concepts [here](https://github.com/gadenbuie/tidyexplain).

Then, let's go through [Jenny Bryans's join cheatsheet](https://stat545.com/join-cheatsheet.html) for examples. 


## Activity

Let's complete [today's worksheet](https://raw.githubusercontent.com/USF-Psych-DataSci/Classroom/master/tutorials/s07_tibble-join-exercise.Rmdd).
